---
title: "Week 2 exercises"
author: "Suvi Vainio"
date: "13 maaliskuuta 2019"
output: html_document
---

```{r setup, include=FALSE}
library(magrittr)
knitr::opts_chunk$set(echo = FALSE)
#setwd("//ad.helsinki.fi/home/s/suvivain/Documents/2019 Survival analysis")
library(eha)
library(survival)
?veteran # for explanation of the variables in the data set
data(veteran) # load the data
str(veteran) # show records of the dat
head(veteran)
```

# Fitting exponential and Weibull model to Veteran data
## 1. Plot a histogram of the survival times corresponding to uncensored observations (veteran$status == 1) as done in Exercise 1.

Median of survival time is 63 days and 90 % of the population has died by the dawn of day number 296.

```{r}

hist(veteran$time[veteran$status == 1]
     , nclass=30, main=''
     , xlab='Survival time (days)'
     , col='gray90')

print("Percentages of the survival")
quantile(veteran$time[veteran$status == 1]
         , probs=seq(0, 1, 0.1))
```

## 2. Compare the Kaplan-Meier estimate of the survival function to (a) Exponential distribution, and (b) Weibull distribution. Use graphical procedure and interpret the results.
Hint: You can obtain the maximum likelihood estimates of the parameters using weibreg() function of eha package.

Weibul distribution seems to fit the data better than exponential distribution.



```{r}
veteran0=veteran[veteran$status == 1,]

analyse1=survfit(Surv(time, status)~1, data=veteran0, conf.type="none")

veteran0.exp0 <- weibreg(formula = Surv(time, status) ~ 1, data=veteran0, shape=1)
lambda0 <- exp(-veteran0.exp0$coeff[1])

print(paste("Lambda-parameter for the exponential distribution is: "
            , round(lambda0, 2), sep="")
      )

veteran0.weibull0 <- weibreg(formula = Surv(time, status) ~ 1, data=veteran0)
b <- exp(veteran0.weibull0$coeff[1])
a <- exp(veteran0.weibull0$coeff[2])

print(paste("Parameters a and b for Weibul-distribution are: "
            , round(a,2), " and ", round(b,2), sep=""))

x1=veteran0[order(veteran0$time),]$time
plot(analyse1)
lines(x1, exp(-lambda0*x1), col="blue", type="l")
lines(x1, exp(-(x1/b)**a), col="red")
legend("topright"
       , legend = c("data", "exponential", "weibul")
       , col=c("black", "blue", "red"), lty=1)
print("Result for the fitted exponential model")
veteran0.exp0
print("Result for the Weibul model")
veteran0.weibull0
```

# Model choice

## 3. Compare the above two models with the likelihood ratio test. Interpret the result. Hint: You can extract the log-likelihood values from the output objects of function weibreg. Use the pchisq function to calculate the p-value (tail probability).

Alternative: You can calculate the likelihood ratio by using the anova command on the output objects from the two regression models using survreg.

Hypotheses for likelihood-ratio test are:

H0: Simple exponential and Weibull model fit the data equally good.
H1: Weibull distribution fits the data better.

Test statistic is -2log(theta0-theta1).

```{r}

print(paste0("Loglikelihood for exponential model: ", veteran0.exp0$loglik[1]))
print(paste0("Loglikelihood for Weibul-model is: ", veteran0.weibull0$loglik[1]))
print(paste0("Test statistic value is: "
             , -2*(chi2.veteran0.exp0-chi2.veteran0.weibul0)[1]
             , " with 1 degree of freedom."
             )
      )
print(paste0("and probability for test statistic to be larger than observed is: "
             , pchisq(-2*(chi2.veteran0.exp0-chi2.veteran0.weibul0)[1]
                      , df=1
                      , lower.tail = F)
             , " thus more complex Weibull-model fits the data better than exponential model."
             )
      )

```

# Simulation
1. Generate 100 random numbers from exponential distribution with mean 0.01 and store it in T.
R's function uses rate as a parameter, mean equals 1/rate.
1. Plot the empirical distribution function.
2. Estimate the rate (stored under obsrate) from the simulated data and overlay the plot of the distribution function 1 − exp(−obsrate  t).
3. Overlay the plot of the true exponential distribution function.
4. Explore the possibilities for different kinds of line and point plots. Vary the plot symbol, line type, line width, and colour. Also, try to give legend in the above graph.

```{r, echo=T}
set.seed(60)
sample0=rexp(100, 1/0.01)
lambdaHat=100/sum(sample0)
print(paste0("Estimated lambda is: ", lambdaHat))
hist(sample0, freq = F, ylim=c(0, 120))
lines(sample0[order(sample0)], 1-lambdaHat*exp(-lambdaHat*sample0[order(sample0)]), col="blue")
lines(sample0[order(sample0)], lambdaHat*exp(-lambdaHat*sample0[order(sample0)]), col="red")
legend("topright", c("Sampled data", "Survival function with lambdaHat", "Probability funtion with lambdaHat")
       , col=c("black", "blue", "red"), lty=1)

```





